Bayesian Sparsification Methods for Deep Complex-valued Networks

### 2020-06-18 paper/paper.tex
2 Overhauled the beginning of the Bayesian section  * more thorough exposition of the transition from Bayesian inference to SGVB  * added useful comments about SGVB versus local reparameterization trick  * cleaned up citations

% segue into why we chose sparse VD rather than anything else
% (sparsity tech large-scale models) gale-elsen-hooker.tex
\citet{gale_state_2019} compare magnitude pruning, $\ell_0$ regularization and Sparse
Variational Dropout (VD) on large-scale models. Their results suggest that VD may achieve
good accuracy-sparsity balance and outperform pruning and $\ell_0$ in deep architectures,
although pruning is preferred for simplicity, stability and speed. They also observe that
VD induces non-uniform sparsity throughout the model, which \citet{he_amc:_2018} have shown
to be essential for superior compression.
% Magnitude pruning induces user-specified sparsity distributions.
%
% L0 and dropout work in the low-to-mid sparsity range, variational dropout consistently
% ranked behind L0 on Transformer, and was bested by magnitude pruning for sparsity
% levels of 80\% and up. Variational dropout consistently produces models on-par or better
% than magnitude pruning on ResNet-50, and l0 fails to produce sparse models at all.
We expect the overall conclusion of \citet{gale_state_2019} to carry over to $\cplx$VNN,
since there is nothing fundamentally different in $\cplx$-valued versions of the methods
they study, except for the need to operate on pairs of parameters.

% with noting that compression and sparsification techniques serve dual purpose as model
% regularization. and can be regarded as a Bayesian network pruning method

% our contribution
Sparse Variational Dropout is a Bayesian Variational Inference method with automatic
parameter relevance determination effect. In this study we extend this technique to
$\cplx$VNN, inspired by the results of \citet{gale_state_2019}, and motivated by seldom
application of Bayesian Inference to $\cplx$-valued networks \citep{popa_complex-valued_2018}
and apparent scarcity of compression methods specific to them. We assess the performance-%
compression trade-off of the extension by conducting a large-scale numerical study on
image classification on MNIST-like and CIFAR10 datasets and music transcription on MusicNet.

The paper is structured as follows. Sec.~\ref{sec:variational_dropout} reviews Bayesian
dropout techniques, and sec.~\ref{sec:c_valued_networks} provides a brief summary of
the inner working of complex-valued networks. The main contribution of this study is
presented in sec.~\ref{sec:c_variational_dropout}, where we outline the details of
different variational approximations and priors and derive the penalty terms in the
evidence lower bound. In sec.~\ref{sec:experiments} we evaluate the sparsification rate,
compare the resulting performance of various $\cplx$-networks proposed in prior work,
and discuss the outcomes.

@@

% segue into bayesian inference
  % \speak{R 1}
  % In the first paragraph of section 2: Although the Bayesian prior over weights can indeed
  % be interpreted as a prior over models, the inclusion of this notation in the first
  % paragraph of section 2 seems unnecessary. Especially considering that the authors refer
  % to weights, and not models, for the remainder of the paper.

Bayesian Inference is a principled framework of reasoning about uncertainty, and in broad
terms Bayesian methods are used to update prior assumptions in accordance with their
likelihood based on empirical evidence into posterior beliefs. Under suitable modelling
and distributional assumptions, Bayesian methods can be used towards model regularization,
e.g. Variational Dropout \citep{kingma_variational_2015}.
% Bayesian perspective on dropout model regularization proposed by Kingma at al. (2015)
% repurposed for sparsification by Molchanov et al. (2017) and Kharitonov et al. (2018)

In general, the core of Bayesian Inference can be summarized as follows: given a prior
distribution $\pi(m)$ on models (hypotheses) $\mathcal{M}$, utilize the empirical evidence
$D$ to update the assumptions by considering the likelihood of the observations under each
$m \in \mathcal{M}$. Models are represented by a parametric family indexed by parameters
$\omega \in \Omega$ and each $m_\omega(\cdot)$ specifies the conditional distribution of
the data $
  D = (z_i)_{i=1}^N
$. The posterior distribution $p(\omega \mid D)$, derived using the Bayes rule $
  p(m \mid D) = \tfrac{p(D \mid m) \pi(m)}{p(D)}
$ with $
  p(D) = \mathbb{E}_{\pi(m)} p(D \mid m)
$, provides useful information about yet unobserved data and model parameters, e.g
classification uncertainty, predictive statistics, and parameter relevance.

Save for the relatively simple cases, the posterior distribution is intractable, and
therefore exactness Bayesian Inference is traded for tractability and scalability of
Variational Bayesian Inference. Instead of deriving the posterior $p(\omega \mid D)$,
this approach postulates an optimization problem for finding an approximation $q$,
which is close to $p(\omega \mid D)$ in terms of a proximity score $\rho$
\begin{equation}  \label{eq:variational-progam}
  q_*
    \in \arg \min_{\theta} \rho\bigl(
      q_\theta(\omega), p(\omega \mid D)
    \bigr)
    \,,
\end{equation}
over a tractable parametric family of distributions on the parameter space $
  \mathcal{Q} = \{Q_\theta(d\omega) \colon \theta \in \Theta\}
$, $\theta$ -- generic variational parameter. The variational approximation class
$\mathcal{Q}$ is chosen so that its members have tractable densities $
  q_\theta(\omega) d\omega
$, can be sampled from and have either tractable $\log$-derivatives $
  \omega \mapsto \nabla_\theta \log q_\theta(\omega)
$ or can be represented as differentiable transformations of non-parametric noise
\citep{williams_simple_1992,kingma_auto-encoding_2014,figurnov_implicit_2019}.
% (reparameterization trick, pathwise gradient) $q_{\theta}(d\omega)$ is a push-forward
% of $p(d\varepsilon)$ by a differentiable map $g(\varepsilon; \theta)$.
% % http://stillbreeze.github.io/REINFORCE-vs-Reparameterization-trick/#fn:1
%
The most common choice for $\rho$ is Kullback-Leibler divergence of density $q$ from $p$,
\begin{equation}  \label{eq:kl-div-def}
  KL(q \| p)
    % = \int \frac{dQ}{dP} \log{\frac{dQ}{dP}} P(d\omega)
    % = \int \frac{q(\omega)}{p(\omega)} \log{\frac{q(\omega)}{p(\omega)}} p(\omega) d\omega
    % = \mathbb{E}_{\omega \sim q}
    %   \log\frac{q(\omega)}{p(\omega)}
    = \mathbb{E}_{\omega \sim q}
      \log{q(\omega)}
    - \mathbb{E}_{\omega \sim q}
      \log{p(\omega \mid D)}
    \,,
\end{equation}
% and, despite it not being a metric and requiring distributions with nested support,
the main reason being that it satisfies the following identity for well-behaved densities:
% $(q, p, \pi)$
\begin{multline}  \label{eq:kl-div-master}
  % \log p(D)
    % = \mathbb{E}_{q(\omega)} \log \frac{p(D, \omega)}{p(\omega \mid D)}
    % = \mathbb{E}_{q(\omega)} \log \frac{
    %   p(D \mid \omega) \pi(\omega)
    % }{p(\omega \mid D)}
    % = \mathbb{E}_{q(\omega)} \log \frac{
    %   p(D \mid \omega) \pi(\omega)
    % }{q(\omega)}
    % + \mathbb{E}_{q(\omega)} \log \frac{q(\omega)}{p(\omega \mid D)}
    % - \mathbb{E}_{q(\omega)} \log \frac{q(\omega)}{\pi(\omega)}
    % = \mathbb{E}_{q(\omega)} \log{p(D \mid \omega)}
    % + \mathbb{E}_{q(\omega)} \log \frac{q(\omega)}{p(\omega \mid D)}
  \log p(D)
    - KL(q \| p(\omega \mid D))
    \\ = \mathbb{E}_{q(\omega)} \log{p(D \mid \omega)}
    - KL(q \| \pi)
    \,.
\end{multline}
This leads to an equivalent problem for \eqref{eq:variational-progam} of maximizing
the \textit{evidence lower bound} (ELBO)
\begin{equation}  \label{eq:elbo_general}
  \mathcal{L}(\theta; \phi, \lambda)
    = - KL(q_{\theta} \| \pi_{\lambda})
      + \mathbb{E}_{\omega \sim q_{\theta}}
        \log p_{\phi}(D \mid \omega)
      % \mathbb{E}_{\omega \sim q_{\theta}}
        % \log\frac{q_\theta(\omega)}{\pi_{\lambda}(\omega)}
  % = \mathbb{E}_{\omega \sim q_{\theta}}
  %     \log{p_{\phi}(D \mid \omega) \pi_{\lambda}(\omega)}
  %   + \mathbb{H}(q_{\theta})
  \,,
\end{equation}
where $\lambda$, $\theta$ and $\phi$ are optimization parameters of the prior $\pi$,
approximation $q$ and the likelihood $p(D\mid \omega)$, respectively. Other objectives
are possible, besides \eqref{eq:elbo_general}, provided $p(\omega \mid D)$ is evaluated
only through $\log p(D \mid \omega)$ and $\log \pi(\omega)$, \citep{ranganath_operator_2018}.

\textit{Stochastic Gradient Variational Bayes} (SGVB), proposed by \citet{kingma_auto-encoding_2014},
replaces the Variational objective \eqref{eq:elbo_general} by its unbiased differentiable
Monte-Carlo estimator, which makes it possible to employ stochastic gradient optimization
methods to solve \eqref{eq:variational-progam}. If $
  \omega \sim q_{\theta}
$ is equivalent in distribution to $
  \omega = g(\varepsilon; \theta)
$ for some differentiable $
  g(\varepsilon; \theta)
$ and non-parametric random variable $
    \varepsilon \sim p_\varepsilon
$, then the SGVB objective is
\begin{equation}  \label{eq:sgvb_objective}
  \widetilde{\mathcal{L}} %(\theta; \phi, \lambda)
    = - KL(q_{\theta} \| \pi_{\lambda})
      % + \frac{N}{\lvert B \rvert} \sum_{i\in B}
        % \log p(z_i \mid \omega_i)b
      + \frac1{L} \sum_{l=1}^L
        \log p_{\phi}(D \mid g(\varepsilon_{l}; \theta))
        % \log p_{\phi}(D \mid \omega_{l})
        %   \Big\vert_{\omega_{l} = g(\varepsilon_{l}; \theta)}
      % =
      % - KL(q_{\theta} \| \pi)
      % + N \mathbb{E}_{B \sim \{D\}_M \otimes q_\theta^M}
        % \hat{\mathbb{E}}_{z,\omega \sim B}
          % \log p(z \mid \omega)
    \,, 
\end{equation}
where $
  (\varepsilon_{l})_{l=1}^L
$ is an iid sample from $p_\varepsilon$, the KL-divergence term replaced by unbiased
differentiable finite sample estimator if it is intractable.
%
For the purposes of this study, we assume that the likelihood satisfies $
  \log p_\phi(D\mid \omega)
    = \sum_i \log p_\phi(z_i \mid \omega)
$, i.e. the observed data $
  D = (z_i)_{i=1}^N
$ is conditionally independent, and consider \eqref{eq:sgvb_objective} with $L=1$
and random mini-batches from $D$:
\begin{equation}  \label{eq:elbo}
  \widetilde{\mathcal{L}} %(\theta; \phi, \lambda)
    = - KL(q_{\theta} \| \pi_{\lambda})
      + \frac{N}{M} \sum_{k=1}^M
        \log p_{\phi}(z_{i_k} \mid g(\varepsilon_k; \theta))
    \,,
\end{equation}
for a random subsample $(z_{i_k})_{k=1}^M$ and iid $\varepsilon_k \sim p_\varepsilon$.
%
% (DSVI) titsias-lazaro.tex \todo{DSVI} \citep{titsias_doubly_2014}

With a special family of posterior approximation, Variational Inference can be used as
a regularization method and as a model sparsification technique.

@@

Dropout can be viewed as a Bayesian model averaging method, where models share parameters
and are weighted equally. Thus, \citet{kingma_variational_2015} consider Dropout, DropConnect
\citep{wan_regularization_2013}, and Gaussian dropout \citep{wang_fast_2013} through the
lens of Bayesian Inference methods and propose \textit{Variational Dropout}. They argue that
the multiplicative noise introduced by these methods induces a distribution equivalent to
a fully factorized variational posterior of the form $
  q_\theta(\omega) = \prod_j q_{\theta}(\omega_j)
$, where $q_{\theta}(\omega_j)$ is $\omega_j = \mu_j \xi_j$ with $
  \xi_j \sim p_\theta(\xi_j)
$ iid from some $p_\theta(\xi)$.

@@

% testing the fx of the threshold
\begin{figure}[!t]
  \centering
  \includegraphics[width=1\columnwidth]{figure__musicnet__threshold__C__DeepConvNet.pdf}
  \caption{%
    Effects of finetunig on performance-compression curves parametrized by $
      \tau\in \{\tfrac{k}2\colon k=-8..+8 \}
    $ for $
      C\in \{\tfrac1{20}, \frac1{200}\}
    $ in \eqref{eq:elbo_with_coef} during sparsification (sec.~\ref{sub:staging}).
  }
  \label{fig:hist__and__threshold__tradeoff}
\end{figure}

>>>>

% segue into why we chose sparse VD rather than anything else
% (sparsity tech large-scale models) gale-elsen-hooker.tex
\citet{gale_state_2019} compare magnitude pruning, $\ell_0$ regularization and Sparse
Variational Dropout (VD) on large-scale models. Their results suggest that VD may achieve
good accuracy-sparsity balance and outperform pruning and $\ell_0$ in deep architectures,
although pruning is preferred for simplicity, stability and speed. They also observe that
VD induces non-uniform sparsity throughout the model, which \citet{he_amc:_2018} have shown
to be essential for superior compression.
% Magnitude pruning induces user-specified sparsity distributions.

% our contribution
Sparse Variational Dropout is a Bayesian Variational Inference method with automatic
parameter relevance determination effect. In this study we extend \emph{Sparse VD} to
$\cplx$VNN, inspired by the results of \citet{gale_state_2019}, and motivated by seldom
application of Bayesian Inference to $\cplx$-valued networks \citep{popa_complex-valued_2018}
and apparent scarcity of compression methods specific to them. We assess the performance-%
compression trade-off of the extension by conducting a large-scale numerical study on
image classification on MNIST-like and CIFAR10 datasets and music transcription on MusicNet.
%
% THIS SHOULD GO TO CONCLUSIONS!
%
% L0 and dropout work in the low-to-mid sparsity range, variational dropout consistently
% ranked behind L0 on Transformer, and was bested by magnitude pruning for sparsity
% levels of 80\% and up. Variational dropout consistently produces models on-par or better
% than magnitude pruning on ResNet-50, and l0 fails to produce sparse models at all.
We expect the overall conclusion of \citet{gale_state_2019} to carry over to $\cplx$VNN,
since there is nothing fundamentally different in $\cplx$-valued versions of the methods
they study, except for the need to operate on pairs of parameters.

The paper is structured as follows. Sec.~\ref{sec:variational_dropout} reviews Variational
Dropout, and sec.~\ref{sec:c_valued_networks} provides a brief summary of the inner
working of complex-valued networks. The main contribution of this study is presented in
sec.~\ref{sec:c_variational_dropout}, where we outline the details of different variational
approximations and priors and derive the penalty terms in the evidence lower bound.
In sec.~\ref{sec:experiments} we evaluate the sparsification rate, compare the resulting
performance of various $\cplx$-networks proposed in prior work, and discuss the outcomes.

@@

% overview
In broad terms Bayesian Inference is a principled framework of reasoning about uncertainty
and a tool to update prior beliefs about model's parameters in accordance with evidence or
empirical data into a posterior distribution. The posterior is useful for inference
regarding unobserved data, predictive statistics, parameter confidence regions, and
model's uncertainty.
% intuitive definition of  of model given data is the same as probability of data under hypothesis (model)
% KL-div min & VI -- \citep{jordan_introduction_1999} approximate Bayesian inference
% SVI -- \citep{hoffman_stochastic_2013} scalable stochastic variant
% SGVB -- \citep{kingma_auto-encoding_2014} differentiable MCMC estimator of the evidence lower bound for SVI

For an observed dataset $
  D = (z_i)_{i=1}^N
$ with parametric statistical model (likelihood) $
  p(D \mid \omega)
    = \prod_i \log p(z_i \mid \omega)
    % $D$ conditional independence given parameters $\omega$
$ the Bayes rule $
  p(\omega \mid D) = \tfrac{p(D \mid \omega) \pi(\omega)}{p(D)}
$ transforms prior hypotheses $\pi(\omega)$ about the unknown distribution of model's
parameters $\omega$ into the posterior distribution $p(\omega \mid D)$.
%
% In general the marginal likelihood of the dataset is intractable $
%   p(D) = \mathbb{E}_{\pi(\omega)} p(D \mid \omega)
% $.
Save for the relatively simple set-ups, either the posterior distribution itself or the
mathematical expectations it is involved in are analytically intractable or impractical
to compute numerically. Thus, Variational Inference (VI), an approximate inference method
proposed by \citet{jordan_introduction_1999}, is used to find an approximation within
a parametric distribution family with densities $q_\theta(\omega)$, that is closest
in terms of Kullback-Leibler (KL) divergence to the true posterior distribution, $
  KL(q_\theta(\omega) \| p(\omega \mid D))
    % = \int \frac{dQ}{dP} \log{\frac{dQ}{dP}} P(d\omega)
    % = \int \frac{q(\omega)}{p(\omega)} \log{\frac{q(\omega)}{p(\omega)}} p(\omega) d\omega
    % = \mathbb{E}_{\omega \sim q}
    %   \log\frac{q(\omega)}{p(\omega)}
    = \mathbb{E}_{\omega \sim q}
      \log \tfrac{q_\theta(\omega)}{p(\omega \mid D)}
$. The approximation family is chosen so that the expectations become tractable.
%
\citet{jordan_introduction_1999} showed that this problem is equivalent to \emph{maximizing}
the \emph{Evidence Lower Bound} (ELBO)
\begin{equation}  \label{eq:elbo_general}
  \mathcal{L}(\theta; \phi, \lambda)
    = - KL(q_{\theta} \| \pi_{\lambda})
      + \sum_{i=1}^N \mathbb{E}_{\omega \sim q_{\theta}}
        \log p_{\phi}(z_i \mid \omega)
      % \mathbb{E}_{\omega \sim q_{\theta}}
        % \log\frac{q_\theta(\omega)}{\pi_{\lambda}(\omega)}
  % = \mathbb{E}_{\omega \sim q_{\theta}}
  %     \log{p_{\phi}(D \mid \omega) \pi_{\lambda}(\omega)}
  %   + \mathbb{H}(q_{\theta})
  \,,
\end{equation}
where $\lambda$ and $\phi$ parameterize the prior and the likelihood, respectively.
%
It is possible to use other objectives, besides \eqref{eq:elbo_general}, provided $
  p(\omega \mid D)
$ is evaluated only through $\log p(D \mid \omega)$ and $\log \pi(\omega)$,
\citep{ranganath_operator_2018}.

To make VI able to handle large-scale datasets \citet{hoffman_stochastic_2013} proposed
\emph{Stochastic Variational Inference}, which uses stochastic gradient optimization over
mini-batches from $D$ to solve \eqref{eq:elbo_general} with exponential families of
distributions. To allow for efficient inference with non-conjugate distributions
in large-scale setting \citet{titsias_doubly_2014} proposed \emph{Doubly Stochastic
Variational Bayes} that uses unbiased Monte-Carlo estimate of the gradient of ELBO
and requires that $q_\theta$ can be efficiently sampled from and have tractable
$\nabla_\theta \log q_\theta(\omega)$.
%
Independently, \citet{kingma_auto-encoding_2014} proposed \emph{Stochastic Gradient
Variational Bayes} estimator for ELBO and its gradient \eqref{eq:elbo_general} with
respect to parameters via an unbiased differentiable Monte-Carlo estimator, provided
the density $q_\theta(\omega)$ is amenable to the \emph{reparameterization trick},
i.e. $
  \omega \sim q_{\theta}
$ is equivalent in distribution to $
  \omega = g(\varepsilon; \theta)
$ for some non-parametric random variable $
    \varepsilon \sim p_\varepsilon
$ and $g(\varepsilon; \theta)$ differentiable with respect to $\theta$.

% HERE
% The Local Reparameterization Trick reduces the variance of SGVB
% even further by sampling independent weight matrices for
% each data-point inside mini-batch. So we need both original SGVB
% (per sample randomness) and local reparameterization (that is an
% efficient and less noisy equivalent to per-sample weight sampling)
%
The SGVB objective is
\begin{equation}  \label{eq:sgvb_objective}
  \widetilde{\mathcal{L}} %(\theta; \phi, \lambda)
    = - KL(q_{\theta} \| \pi_{\lambda})
      % + \frac{N}{\lvert B \rvert} \sum_{i\in B}
        % \log p(z_i \mid \omega_i)b
      + \frac1{L} \sum_{i=1}^N \sum_{l=1}^L
        \log p_{\phi}(z_i \mid g(\varepsilon_{il}; \theta))
        % \log p_{\phi}(D \mid \omega_{l})
        %   \Big\vert_{\omega_{l} = g(\varepsilon_{l}; \theta)}
      % =
      % - KL(q_{\theta} \| \pi)
      % + N \mathbb{E}_{B \sim \{D\}_M \otimes q_\theta^M}
        % \hat{\mathbb{E}}_{z,\omega \sim B}
          % \log p(z \mid \omega)
    \,,
\end{equation}
where $
  (\varepsilon_{l})_{l=1}^L
$ is an iid sample from $p_\varepsilon$. The KL divergence term can replaced by an
unbiased differentiable finite sample estimator if it is intractable.

For the purposes of this study, we follow \citet{kingma_auto-encoding_2014} and use
\eqref{eq:sgvb_objective} with random batches of size $M$ from $D$ and $L=1$ $\varepsilon$
samples per datapoint in the mini-batch:
\begin{equation}  \label{eq:elbo}
  \widetilde{\mathcal{L}} %(\theta; \phi, \lambda)
    = - KL(q_{\theta} \| \pi_{\lambda})
      + \frac{N}{M} \sum_{k=1}^M
        \log p_{\phi}(z_{i_k} \mid g(\varepsilon_k; \theta))
    \,,
\end{equation}  % \citet{kingma_auto-encoding_2014}
for a random subsample $(z_{i_k})_{k=1}^M$ and iid $\varepsilon_k \sim p_\varepsilon$.

\medskip
With a special posterior approximation family and suitable prior assumptions, Variational
Inference can be used as model regularization and sparsification technique.

@@

% VD is a Bayesian perspective on dropout regularization \citep{hinton_improving_2012} offered
% by \citet{kingma_variational_2015} and repurposed for sparsification by \citet{molchanov_variational_2017}.
%
Dropout can be viewed as a Bayesian model averaging method, where models share parameters
and are weighted equally. Thus, \citet{kingma_variational_2015} consider Dropout, DropConnect
\citep{wan_regularization_2013}, and Gaussian dropout \citep{wang_fast_2013} through the
lens of Bayesian Inference methods and propose \textit{Variational Dropout}. They argue that
the multiplicative noise introduced by these methods induces a distribution equivalent to
a fully factorized variational posterior of the form $
  q_\theta(\omega) = \prod_j q_{\theta}(\omega_j)
$, where $q_{\theta}(\omega_j)$ is $\omega_j = \mu_j \xi_j$ with $
  \xi_j \sim p_\theta(\xi_j)
$ iid from some $p_\theta(\xi)$.

@@

% testing the fx of the threshold
\begin{figure}[!t]
  \centering
  \includegraphics[width=1\columnwidth]{figure__musicnet__threshold__C__DeepConvNet.pdf}
  \caption{%
    Effects of finetunig on performance-compression curves parameterized by $
      \tau\in \{\tfrac{k}2\colon k=-8..+8 \}
    $ for $
      C\in \{\tfrac1{20}, \frac1{200}\}
    $ in \eqref{eq:elbo_with_coef} during sparsification (sec.~\ref{sub:staging}).
  }
  \label{fig:hist__and__threshold__tradeoff}
\end{figure}
